Computational methods in landscape ecology

### 2024-07-02 Sections/05-simulations.qmd
Update simulations section

## Landscapes generators

Landscape generators are typically used as null hypotheses, baselines, or scenarios in landscape ecology analyses because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler as ecological simulations models [@Hess2020].
Landscape generators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013].

Pattern-based landscape generators simulate landscape patterns without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or include hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics and being able to simulate hierarchical landscapes, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or Binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities, e.g., based on least-cost networks [@Etherington2024].
In order to ensure realistic landscapes, several approaches exist that use real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022] to simulate neutral landscapes.
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], or Julia (https://github.com/EcoJulia/NeutralLandscapes.jl). -->
<!-- QRULE -->

Contrastingly, processes-based landscape generators explicitly include hypothesized pattern creating processes [@Langhammer2019; @Salecker2019].
<!-- are able to simulate structurally more realistic landscapes by including -->

<!--
Cellular automata
[G-RaFFE](https://doi.org/10.1371/journal.pone.0064968)
[PioLaG](https://doi.org/10.1007/s10980-020-01066-w)
-->

## IBMs /Coupled lattice

@@

<!-- 
[RangeShifter](https://doi.org/10.1111/ecog.05687)
[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
[SDM past present future](https://doi.org/10.1111/jbi.14617)
[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
[HexSim modeling environment](https://doi.org/10.1007/s10980-017-0605-9)
[Digital twins?](https://doi.org/10.1016/j.tree.2023.04.010)
-->

>>>>

## Landscapes simulators

Landscape simulators are typically used as null hypotheses, baselines, or scenarios because they allow to control certain aspects of the landscape [@Li2004; @Wang2008] and are generally simpler than ecological simulations models [@Hess2020].
Landscape simulators can be classified into two major categories, namely pattern-based and process-based approaches [@Peer2013; Langhammer2019].

Pattern-based approaches simulate landscapes without assuming any underlying abiotic or biotic processes [i.e., neutral landscape model\; @Gardner1987].
The earliest neutral models are based on percolation theory and assign LULC types randomly to cells in the landscape[@Gardner1987] or hierarchical models that consider different spatial scales while assigning cell values [@ONeill1992].
Landscapes characterized by continuous values can be simulated by fractal models, such as Brownian motion [@Palmer1992].
Borrowing from computer graphics, more recent neutral landscape models make use of spectral synthesis [e.g., Perlin noise\; @Etherington2022] or binary space partitioning [@Etherington2022a].
Neutral landscape models are also able to simulate landscapes dominated by anthropogenic activities using based on least-cost networks [@Etherington2024].
In order to ensure realistic neutral landscapes several approaches exist that use comparisons with real landscapes [@Inkoom2017] or are based on target values optimization [@VanStrien2016; @Justeau-Allaire2022].
<!-- In order to make multiple neutral landscape model algorithms accessible, several software libraries are available in prominent programming languages including Python [NLMpy\; @Etherington2015], R [NLMR\; @Sciaini2018], Julia (https://github.com/EcoJulia/NeutralLandscapes.jl), or GUI software (GradientLand, QRULE) -->

Contrastingly, processes-based approaches explicitly include abiotic or biotic pattern-forming processes [i.e., landscape generators\; @Langhammer2019; @Salecker2019].
Several earlier landscape generators are based on cellular automata models and are able to simulate urban growth [@Clarke1998] or deforestation [@Soares-Filho2002].
More recent landscape generators allow to simulate patchy landscapes based on Gibbs processes [@Gaucherel2006], deforestation based on road and agricultural fields access [@Peer2013], agricultural areas dominated by smallholders [@Salecker2019], or vegetation surround watering points in semi-arid savanna rangelands [combining pattern- and process-bases approaches\; @Hess2020].

## Individual-based models

Individual-based models (or agent-based models) simulate discrete entities that are described by attributes and behaviors and patterns emerge from bottom-up interactions of individuals with each other and their environment [@DeAngelis2014].
In landscape ecology individual-based models are increasingly used to model especially social-ecological systems [e.g, @Carauta2018; @Dislich2018; @Gonzalez-Redin2019], but also disturbances [e.g., @Seidl2017a; @Spies2017; @Kamimura2019] or connectivity [e.g., @Day2020].
In order to facilitate future model developments, recent progress includes adaptable modeling frameworks [@Schumaker2018] or reusable building block [@Berger2024].
Furthermore, hybrid approaches that couple or even integrated different model types with individual-based models could benefit from rich development history or facilitate corresponding strengths [@OSullivan2016], e.g., linking an individual-based and ecosystem model [@Polhill2011].

@@

<!--
[Digital twins](https://doi.org/10.1016/j.tree.2023.04.010) 
[Rural ecological landscape control](https://doi.org/10.1186/s12302-024-00888-8)
[Urban Expansion and Vegetation Coverage](https://doi.org/10.3389/fpls.2022.840471)
[Digital Twins in agriculture](https://doi.org/10.1016/j.cosust.2022.101252)
[Future landscape visualization](https://doi.org/10.1093/jcde/qwac032)
[Digital Twin of the Agricultural Landscape](https://doi.org/10.14627/537690019)
-->

<!--
[Marine SDMS](https://doi.org/10.1016/j.ecolmodel.2019.108837)
[SDM past present future](https://doi.org/10.1111/jbi.14617)
[Simulations in landscape genetics](https://doi.org/10.1111/j.1365-294X.2010.04678.x)
-->
